Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.

### 2025-01-17 02-antecedentes.qmd
ajustes antecedentes

A shock like the COVID-19 pandemic posed a considerable threat to economically vulnerable groups and those in precarious labor market positions[@ministerio_de_desarrollo_social_y_familia_resultados_2022;@instituto_nacional_de_estadisticas_resultados_2022,]. In 2015, the poverty rate was 11.2%, decreasing to 8.5% by 2017, with the 10/10 ratio at 31.6. In 2020, during the COVID-19 pandemic, poverty increased to 10.7%, and the 10/10 ratio surged to 301.7, while unemployment reached 10.5%, the highest level in a decade. By 2022, poverty declined to 6.5%, the 10/10 ratio improved to 61.2, and unemployment decreased to 7.9%, rising slightly to 8.7% in 2023.

@@

Besides proposing a general decrease over time in the justification of inequalities, a second hypothesis refers to specific aspects of the pandemic. As some political measures from the state included several social benefits (as delivering food boxes, free vaccination, bonuses, among others), we argue that being recipient of such benefits could affect attitudes towards the redistributive role of the state as equalizer opportunities to face risks. As a consequence, preferences for market justice would be reduced:

$H_2$: Recipient of state benefits show lower market justice preferences

In the same line, we expect that those who give priority to take care of the health of the population instead of the country economy would exhibit larger market justice preferences:

$H_3$: Those prioritizing health instead of the economic situation show higher market justice preferences

@@

Although there are few studies relating meritocratic beliefs and inequality justification, from a broader perspective it is possible to find research which incorporates the support of different distributive principles with regard to education. For instance, in an experiment by @igliozziFairShareEffects2024, they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education. They find that a larger system justification  [@jost_psychology_2003] is linked to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school'). In the same line, @leeFairnessPerceptionsEducational2023 show that performance assessments (in what they call “neoliberal orientations”) predicted people's fair perceptions of educational inequality.

@@

Empirical cross-sectional evidence from Chile reveals two significant trends: low public support for the perception of meritocracy and limited justification for unequal access to social services such as education and health. For instance, @castillo_clivajes_2013 analyzed data from the International Social Survey Programme (ISSP) in 1999 and 2009, highlighting consistently low support for inequality based on resource allocation or individual merit. Similarly, @Castilloetal2019 demonstrated widespread skepticism among Chileans regarding the meritocratic system, with many disagreeing with the notion that rewards are fairly distributed based on talent and effort. In line with this evidence, we propose that

>>>>

In Chile, the COVID-19 pandemic posed a significant threat to economically vulnerable groups and those in precarious labor market positions [@ministerio_de_desarrollo_social_y_familia_resultados_2022; @instituto_nacional_de_estadisticas_resultados_2022]. The poverty rate, which was 11.2% in 2015, decreased to 8.5% by 2017, with a 10/10 ratio of 31.6. However, during the pandemic in 2020, poverty rose to 10.7%, the 10/10 ratio spiked to 301.7, and unemployment reached a decade-high of 10.5%. By 2022, poverty declined to 6.5%, the 10/10 ratio improved to 61.2, and unemployment decreased to 7.9%, with a slight rise to 8.7% in 2023.

@@

Besides proposing a general decrease over time in the justification of inequalities, a second hypothesis refers to specific aspects of the pandemic. As some political measures from the state included several social benefits (as delivering food boxes, free vaccination, direct transfers, among others), we argue that being recipient of such benefits could affect attitudes towards the redistributive role of the state as equalizer opportunities to face risks. As a consequence, preferences for educational inequalities would be reduced:

$H_2$: Recipient of state benefits show lower justification of educational inequalities

In the same line, we expect that those who give priority to take care of the health of the population instead of the country economy would exhibit larger preferences for educational inequalities:

$H_3$: Those prioritizing health instead of the economic situation show higher justification of educational inequalities

@@

Although there are few studies relating meritocratic beliefs and inequality justification, from a broader perspective it is possible to find research which incorporates the support of different distributive principles with regard to education. For instance, in an experiment by @igliozziFairShareEffects2024, they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education. They find that a larger system justification  [@jost_psychology_2003] is linked to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school'). In the same line, @leeFairnessPerceptionsEducational2023 show that performance assessments (in what they call “neoliberal orientations”) predicted perceived fairness of educational inequality.

@@

Empirical cross-sectional evidence from Chile reveals two significant trends: low public support for the perception of meritocracy and limited justification for unequal access to social services such as education and health. For instance, @castillo_clivajes_2013 analyzed data from the International Social Survey Programme (ISSP) in 1999 and 2009, highlighting consistently low support for inequality based on resource allocation or individual merit. Similarly, @Castilloetal2019 demonstrated widespread skepticism among Chileans regarding the meritocratic system, with many disagreeing with the notion that rewards are fairly distributed based on talent and effort. In line with this evidence, we propose that:
